---
title: "Introduction to MBRarefy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignettes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, include=FALSE}
set.seed(123456)
```

# Overview

The `MBRarefy` package provides a multi-bin rarefying workflow for alpha diversity association analysis under heterogeneous library sizes in high-throughput count-profile datasets, including immune-repertoire and microbiome sequencing data.

This package is especially useful when:

- your dataset contains highly heterogeneous library sizes across samples;
- you want to reduce residual library-size dependence (confounding effects from library size) in downstream alpha diversity association analysis;
- you want to use repeated rarefying to stabilize diversity estimates across subsampling replicates;
- you want to select library-size bin cutpoints in a data-adaptive and reproducible way rather than using ad-hoc thresholds;
- you want to combine bin-specific association results through cross-bin meta-analysis.


## Required packages

`MBRarefy` relies on the following Comprehensive R Archive Network (CRAN) packages:

- doParallel
- parallel
- foreach
- doParallel
- utils
- ggplot2
- GAReg

These packages are all available on CRAN.

## Installation

Install from the local file (the .tar.gz file), which you can download from https://github.com/mli171/MBRarefy. 

```{r eval=FALSE, echo=TRUE}
devtools::install_local("MBRarefy_0.1.0.tar.gz", dependencies = TRUE)
```

Install from CRAN via

```{r eval=FALSE, echo=TRUE}
install.packages("MBRarefy")
```

A github developer version is also available at https://github.com/mli171/MBRarefy via R package `pak` via

```{r eval=FALSE, echo=TRUE}
pak::pak("mli171/MBRarefy")
```


## Open the Vignette in R 

```{r eval=FALSE, echo=TRUE}
browseVignettes("MBRarefy")
```

or

```{r eval=FALSE, echo=TRUE}
vignette("MBRarefy_vignette", package = "MBRarefy")
```

Then call the library via 

```{r librarysetup, echo=TRUE}
library(MBRarefy)
library(GAReg)
library(changepointGA)
```

## Key Features

- **Multi-bin rarefying:** Stratifies samples by library size and rarefies samples within each bin to a bin-specific depth.
- **Data-adaptive cutpoint selection:** Selects library-size bin cutpoints by minimizing residual within-bin association between alpha diversity and original library size.
- **Fixed-\(K\) and varying-\(K\) modes:** Supports user-specified bin complexity through fixed-\(K\) selection and data-adaptive bin complexity through varying-\(K\) selection.
- **Alpha diversity metrics:** Supports observed richness, Shannon entropy, Simpson/Gini-Simpson index, Chao1, bias-corrected Chao1, and Pielou evenness.
- **Repeated rarefying:** Supports repeated subsampling to stabilize alpha-diversity estimates across rarefying replicates.
- **Parallel execution:** Supports multi-core computation through `foreach` and `doParallel`.
- **Cross-bin meta-analysis:** Combines bin-specific association results using equal, sample-size, or inverse-variance weighting.
- **File-based sample-wise processing:** Reads sample count files individually to reduce cohort-level memory requirements.
- **Residual library-size diagnostics:** Tests whether bin-anchored alpha diversity remains detectably associated with original library size after multi-bin rarefying.


## Main Functions

- `multibin.rarefy.diversity()`: Runs repeated rarefying over a user-specified depth grid and computes alpha diversity metrics for each eligible sample.
- `rarefy.alpha()`: Computes alpha diversity metrics after rarefying for a single sample or count profile.
- `get_alpha_metric_matrix()`: Aggregates replicate-resolved rarefying results into sample-by-depth alpha diversity matrices.
- `fixBinRegObj()`: Objective function for fixed-\(K\) GA-based library-size cutpoint selection.
- `varBinRegObj()`: Objective function for varying-\(K\) GA-based library-size cutpoint selection.
- `multibin.meta.test.alpha()`: Performs bin-wise alpha diversity association tests and combines bin-specific results by cross-bin meta-analysis.

Cutpoint optimization is carried out using `GAReg::gareg_knots()` with `fixBinRegObj()` or `varBinRegObj()` supplied as the objective function.


# Multi-bin Workflow for TCR and Microbiome Alpha Diversity Association Analysis

The package vignettes illustrate the full \texttt{MBRarefy} workflow using two real datasets from distinct application domains. First, we use the TCR immunosequencing dataset from Emerson et al. (2017), which profiled healthy donors with known Cytomegalovirus (CMV) serostatus, to demonstrate alpha-diversity association testing in immune repertoire data. Second, we include a microbiome example using the longitudinal wild baboon gut 16S rRNA dataset available via \texttt{microbiomeDataSets::baboongut()}, which contains 16,234 microbiome profiles from 585 baboons collected over 14 years, together with rich sample metadata. These two examples highlight how the same grid-based profiling, GA-based cutpoint selection, and downstream multibin inference can be applied across immune repertoire and microbiome/count-based profiling studies under heterogeneous library sizes.

The main steps are:

1. **Prepare input files and metadata.** Each sample is represented by a feature-count file, and sample metadata are aligned by sample identifiers.
2. **Compute rarefaction profiles.** `multibin.rarefy.diversity()` rarefies each eligible sample over the specified depth grid and computes alpha diversity metrics.
3. **Summarize replicate-resolved results.** `get_alpha_metric_matrix()` averages over repeated rarefaction replicates and returns a sample-by-depth alpha diversity matrix.
4. **Select library-size cutpoints.** `GAReg::gareg_knots()` is used with `fixBinRegObj()` for fixed-\(K\) selection or `varBinRegObj()` for varying-\(K\) selection.
5. **Extract bin-anchored alpha diversity.** Samples are assigned to library-size bins, and diversity is evaluated at each bin’s lower rarefying depth.
6. **Run association tests and meta-analysis.** `multibin.meta.test.alpha()` performs bin-wise association testing and combines evidence across bins using equal, sample-size, or inverse-variance weighting.
7. **Check residual library-size dependence.** The bin-anchored diversity values are tested against original library size as a diagnostic before interpreting biological or ecological associations.

The TCR example uses observed clonotype richness as the alpha diversity metric and evaluates associations with age, biological sex, and CMV serostatus. The microbiome example uses observed ASV richness and evaluates associations with age, sex, and group size.

## TCR Alpha Diversity Association Analysis

The TCR dataset used in this analysis originates from Emerson et al. (2017), a study designed to identify pathogen exposure signatures from shared TCR sequences. Specifically, the data includes immunosequencing profiles from 786 healthy bone marrow donors, each with known cytomegalovirus (CMV) serostatus. For this analysis, we focus on Cohort 1, which contains 666 samples. Among these, three samples have library sizes below $10^5$, and are filtered out accordingly in downstream analyses for quality control consideration. This dataset enables investigation of the relationship between TCR diversity and CMV exposure, providing a real-world case study for evaluating alpha diversity under the proposed multi-bin rarefying framework.

Note that, due to the large size of TCR CMV data set, only the calculated alpha-diversity results are provided in this package. If users want to have the full data-processing steps replicated, it will be necessary to download the CMV TCR immunosequencing profiles directly from immuneACCESS sites at https://doi.org/10.21417/B7001Z.

1. Prepare Input: Per-sample count tables;

For each TCR repertoire profile, we extract and retain only the essential information: the nucleotide sequence (composed of A, T, G, C) and its corresponding read count. These cleaned profiles are saved as plain text (.txt) files, with one file per sample, and stored within a designated input directory.

```{r eval=FALSE, echo=TRUE}
InputDataDir = "data/"
```

Each processed file should contain at least two columns:

- A sequence identifier column (e.g., "seq"), containing nucleotide strings.
- A count column (e.g., "count"), representing the observed frequency of each sequence.

These .txt files are then used as input to the rarefying pipeline, allowing for scalable diversity estimation while minimizing memory usage by streaming from disk instead of loading all data into memory.

2. Compute repeated rarefying profiles over a depth grid.

To evaluate the confounding relationship between the estimated alpha diversities and the original library sizes across a range of rarefying depths, `MBRarefy` performs multi-bin rarefying over a user-specified grid of rarefying depths using `multibin.rarefy.diversity()`, repeating the process across multiple replicates (`nRep`) and optionally parallelizing the computation using the `doParallel` package. 

For input, each sample is represented by a compact count profile saved as a plain-text file (.txt) in a user-specified directory (`InputDataDir` above). Each file contains the unique sequence identifiers (e.g., ASVs/OTUs for microbiome data or CDR3 nucleotide sequences for TCR data) and their corresponding counts. This lightweight format avoids dependence on platform-specific raw outputs while preserving the information required for rarefying-based diversity estimation.

```{r eval=FALSE, echo=TRUE}
SeqVar   <- "seq"
CountVar <- "count"

depths   <- seq(from=1e5, to=1e7, by=1e5)

alphamethods  <- c(
  "unique_seq_alpha",      # Number of unique taxa (Richness)
  "shannon_alpha",         # Shannon entropy
  "gini_simpson_alpha",    # Gini-Simpson index
  "chao1_alpha",           # Chao1 richness estimator
  "chao1_bc_alpha",        # Bias-corrected Chao1
  "pielou_alpha"           # Pielou's evenness
)

tim1 <- Sys.time()
raw.multibin.alpha.res <- multibin.rarefy.diversity(
  InputDataDir=InputDataDir,
  CountVar=CountVar,
  div.measure="alpha",
  depths=depths,
  methods=alphamethods,
  nRep=10,
  parallel=TRUE,
  nCore=10)
tim2 <- Sys.time()

tim2 - tim1
```

The example nested list structure of the output `raw.multibin.alpha.res` can be checked by `str()` or be found through the R help documentation of `multibin.rarefy.diversity()`. Also note, this function can be used for classical (single-depth) rarefying approaches by specifying only one rarefying depth in the `depths` vector and setting `nRep = 1`. This allows you to obtain diversity metrics from a standard single rarefied dataset using the same interface, without engaging in multi-bin processing.

3. Data-adaptive library-size binning via optimal thresholds selection.

A key step in multibin rarefying is defining library size cutpoints that partition samples into bins for within-bin rarefying and inference. `MBRarefy` provides an automated, data-adaptive procedure to select these cutpoints from the observed library-size distribution, reducing ad hoc choices based on visual inspection or fixed rules.

Given a user-specified rarefying depth grid (`depths` from above) and a target number of bins $K$ (or a complexity penalty), the function `optimalbin()` searches for a common set of ordered cutpoints that yields bins with adequate sample size and stable estimation. Cutpoints are chosen to minimize residual within-bin association between rarefying-based alpha diversity and original library size evaluated at bin anchor depths, optionally adjusting for nuisance covariates (e.g., batch or sequencing run). This criterion targets bins in which alpha-diversity estimates are least confounded by library size, improving comparability across samples while retaining more information than single-depth rarefying. Details can be found in Li (2026).

The optimization is performed using a genetic algorithm (GA) constraint-aware search routine that enforces ordered cutpoints, minimum bin size, and optional minimum spacing on the depth grid. `MBRarefy` returns the selected cutpoints, bin assignments, and diagnostics (objective values and bin-level summaries) to support reproducible reporting and sensitivity analyses.

Next, load the pre-processed alpha diversity data generated by the code chunk above and included with the package.

```{r}
data(CMV1)
# load("../data/CMV1.rda")

dataPheno              = CMV1$dataPheno
raw.multibin.alpha.res = CMV1$alpha.res
depths                 = CMV1$depths
alphamethods           = CMV1$methods
```

Here, the total library size (`totalReads`) is calculated for each sample by summing the count column. The names are aligned with the sample names in the alpha diversity matrix. The corresponding phenotype metadata (`dataPheno`) is loaded and aligned with the same sample ordering. The total read counts are appended to the phenotype table (`dataPheno`) for further analysis.

We then summarize replicates. After performing multiple rarefying replicates, the function `get_alpha_metric_matrix()` is used to aggregate alpha diversity metrics for each sample and rarefying depth. For each diversity measure, it returns a matrix where rows correspond to samples and columns correspond to rarefying depths. Each entry in the matrix represents the average value of the corresponding metric across all replicates. Here we pick "the number of unique sequences", also known as richness, as the illustrative example.

```{r}
alpha.mats <- get_alpha_metric_matrix(raw.multibin.alpha.res, metrics=alphamethods)
USC <- alpha.mats$unique_seq_alpha
dataPheno <- dataPheno[match(rownames(USC), rownames(dataPheno)), ]
```

In general, the overall rarefying technique will be applied to reduce the varying library size, so called confounding `artifacts`, due to the unavoidable sequencing experiment bias. We select the rarefying level to be 1e5 (minimum library size from all samples) and 1e6 (the elbow point from all samples' rarefaction curve), respectively and show the results below.

```{r}
overallRarefycol <- "RarefyTo1e+05"
myfit <- lm(USC[,overallRarefycol] ~ dataPheno$totalReads)
Overallpvalue <- summary(myfit)$coefficients[2,"Pr(>|t|)"]
Overallpvalue

overallRarefycol <- "RarefyTo1e+06"
myfit <- lm(USC[,overallRarefycol] ~ dataPheno$totalReads)
Overallpvalue <- summary(myfit)$coefficients[2,"Pr(>|t|)"]
Overallpvalue

overallRarefycol <- "RarefyTo1800000"
myfit <- lm(USC[,overallRarefycol] ~ dataPheno$totalReads)
Overallpvalue <- summary(myfit)$coefficients[2,"Pr(>|t|)"]
Overallpvalue
```

These diagnostics indicate that overall rarefying can leave detectable residual association between observed richness and original library size. The inference results by testing the association between alpha diversity and phenotypes are questionable. 

Now we describe the multi-bin rarefying approach with optimal bin thresholds selection.

Using a grid of $M=100$ rarefying levels spanning from 1e5 to 1e7 from `depths` above, we form a monotone grid set for the candidate optimal bin thresholding values. Before the multibin association analysis, we choose an optimal binning of $\boldsymbol{L}_{\text{grid}}$ that minimizes residual dependence between the original library size and alpha diversity within bins (Li et al., 2026). Given a rarefying depth grid of length $M=100$ and the pre-computed alpha diversity matrix, we consider two real-world application scenarios: (i) the number of cutting thresholds is specified, or (ii) it must be estimated. For each scenario, we formulate a dedicated objective function and optimize it using the R package `GAReg` (Li et al., 2026). The proposed objective functions, `fixBinRegObj()` and `varyBinRegObj()`, are designed to meet our specific needs and are included in our `MBRarefy` package.

- Model 1: With user-specified number of bins (``fixed-K'');

```{r fixnumbbin, fig.width=10, fig.height=5}
Lorig = dataPheno$totalReads
Lgrid = depths
Y     = as.matrix(USC)

fix_fit <- GAReg::gareg_knots(
  y         = rep(0, length(Lgrid)),
  x         = Lgrid,
  ObjFunc   = fixBinRegObj,
  minDist   = 1L,
  gaMethod  = "cptga",
  cptgactrl = cptgaControl(
    popSize      = 400,
    maxgen       = 100000,
    pchangepoint = 0.3
  ),
  fixedknots   = 5L,
  Y            = Y,
  Lorig        = Lorig,
  Z            = NULL,
  weight       = "count",
  min_subjects = 20L,
  use_logL     = TRUE
)
summary(fix_fit)

best_idx_fix <- fix_fit@bestsol
best_knots_fix <- Lgrid[best_idx_fix]
depths_op_fix <- c(min(Lgrid), best_knots_fix)

res_fix <- plotMBRarefy(
  Lorig = Lorig,
  Lgrid = Lgrid,
  Y = Y,
  fit = fix_fit,
  y_mode = "bin_lower_bound",
  title_prefix = "Select cutpoints to reduce depth-diversity dependence (Fixed-K)",
  y_label = "Observed richness at bin lower bound",
  show_missing_as_rug = FALSE
)

res_fix$plot
```

- Model 2: With estimated number of bins (``vary-K'');

```{r varynumbbin, fig.width=10, fig.height=7}
var_fit <- GAReg::gareg_knots(
  y         = rep(0, length(Lgrid)),
  x         = Lgrid,
  ObjFunc   = varBinRegObj,
  gaMethod  = "cptga",
  minDist   = 3L,
  cptgactrl = cptgaControl(
    popSize      = 400,
    maxgen       = 100000,
    pchangepoint = 0.3
  ),
  Y            = Y,
  Lorig        = Lorig,
  Z            = NULL,
  weight       = "count",
  min_subjects = 20L,
  use_logL     = TRUE
)
summary(var_fit)

best_idx_var <- var_fit@bestsol
best_knots_var <- Lgrid[best_idx_var]
depths_op_var <- c(min(Lgrid), best_knots_var)

res_var <- plotMBRarefy(
  Lorig = Lorig,
  Lgrid = Lgrid,
  Y = Y,
  fit = var_fit,
  y_mode = "bin_lower_bound",
  title_prefix = "Select cutpoints to reduce depth-diversity dependence (Vary-K)",
  y_label = "Observed richness at bin lower bound",
  show_missing_as_rug = FALSE
)

res_var$plot
```

Next, we use detected as the optimal bin thresholds and continue the illustration of multibin rarefying algorithm. Since the detected bin threshold parameters are only interior knots, we need to add-in the minimum of sample library size ($1e5$) as the lower boundary point.

```{r}
depths_op <- depths_op_fix
BinCuts <- c(depths_op, Inf)
```


4. Multi-bin rarefying and alpha-diversity extraction given selected optimal thresholds.

After selecting the optimal cutpoints $\hat{\boldsymbol c}$, the package (i) assigns samples to bins based on their library sizes, (ii) determines a bin-specific rarefying depth $m_k$ as the minimum library size within each bin, and (iii) rarefies all samples in bin $k$ to $m_k$ (subsampling without replacement). Alpha-diversity metrics are computed on the rarefied profiles and can be summarized within bins and across bins. When $R>1$, \texttt{MBRarefy} repeats the rarefying step $R$ times per bin and reports the mean (and optional standard deviation) of the diversity estimates across replicates, improving stability when counts are sparse. The function returns the selected cutpoints, bin assignments, depths $\{m_k\}$, per-sample diversity estimates, and diagnostic summaries to support reproducible reporting and sensitivity analyses.

Given the overall distribution of library sizes (library sizes) ${L_i}$, $i=1,\ldots,n$, and the selected $K-1$ optimal cutpoints, $\hat{c}_{k}$, $k=1,\ldots,K-1$, the TCR profiles are divided into $K$ bins. The choice of threshold values for binning should reflect the observed range and distribution of all $L_i$ across samples. Let:
$$min(L_{i})=c_{0}\leq c_{1}<\ldots<c_{K-1}\leq c_{K}=max(L_{i})$$
Each bin is defined by $[c_{k-1}, c_k)$, and within each bin, multiple rarefying replicates are performed to reduce the stochastic variation introduced by sub-sampling. The alpha diversity metrics are then computed as the average of multiple rarefying realizations per sample per bin. As mentioned in Li et al. (2024), the multi-bin rarefying method improves the robustness and interpretability of alpha diversity estimation in TCR sequencing data, especially in the presence of high variability in library size across samples.

```{r}
USC_mat <- USC[,paste0("RarefyTo", depths_op)]
```

5. Cross-bin association testing and Meta-analysis.

In this step, we perform statistical testing to assess the association between alpha diversity (specifically, the number of unique TCR sequences) and external covariates such as library size, gender, and CMV disease status. The multi-bin rarefying framework allows for more robust meta-analysis by accounting for variability across library size strata. Results are aggregated using three different weighting strategies across bins: equal weighting, sample-size weighting, and inverse-variance weighting. These approaches are detailed in Li et al. (2024). The results of all three weighting methods are stored in the returned R object. In this analysis, we use the inverse-variance weighting method as an example, which is recommended by Li et al. (2024). The results indicate no significant association between library size and alpha diversity, suggesting that the multi-bin rarefying strategy has effectively mitigated the confounding influence of library size. Consequently, downstream associations between alpha diversity and other phenotype variables can be interpreted with greater validity and robustness.

Test association between alpha diversity and original library sizes. A non-significant residual library-size diagnostic with p-value$>0.05$ suggests reduced detectable residual library-size dependence under the fitted diagnostic model.


```{r}
res.test.libsize <- multibin.meta.test.alpha(
  y.alpha.mat = USC_mat,
  xs          = dataPheno$totalReads,
  totalReads  = dataPheno$totalReads,
  BinCuts     = BinCuts
)
# res.test.libsize$Multi.bin.Equal
# res.test.libsize$Multi.bin.SSW
res.test.libsize$Multi.bin.IVW
```

Test association between alpha diversity and age The obtained $<0.05$ p-value still show the significant association between estimated alpha diversity and the age variable.

```{r}
Age <- as.numeric(dataPheno$Age)
res.test.Age <- multibin.meta.test.alpha(
  y.alpha.mat = USC_mat,
  xs          = Age,
  totalReads  = Lorig,
  BinCuts     = BinCuts
)
# res.test.Age$Multi.bin.Equal
# res.test.Age$Multi.bin.SSW
res.test.Age$Multi.bin.IVW
```

Test association between alpha diversity and gender. The obtained $<0.05$ p-value still show the significant association between estimated alpha diversity and the gender variable.

```{r}
sex_x <- as.numeric(as.factor(dataPheno$`Biological Sex`)) - 1
res.test.sex <- multibin.meta.test.alpha(
  y.alpha.mat = USC_mat,
  xs          = sex_x,
  totalReads  = Lorig,
  BinCuts     = BinCuts
)
# res.test.sex$Multi.bin.Equal
# res.test.sex$Multi.bin.SSW
res.test.sex$Multi.bin.IVW
```

Test association between alpha diversity and the observed CMV disease status. The obtained $<0.05$ p-value still show the significant association between estimated alpha diversity and the of-interested clinical variable, virus diseases status.

```{r}
CMV <- as.numeric(as.factor(dataPheno$`Virus Diseases`)) - 1
res.test.CMV <- multibin.meta.test.alpha(
  y.alpha.mat = USC_mat,
  xs          = CMV,
  totalReads  = Lorig,
  BinCuts     = BinCuts
)
# res.test.CMV$Multi.bin.Equal
# res.test.CMV$Multi.bin.SSW
res.test.CMV$Multi.bin.IVW
```


## Microbiome Alpha Diversity Association Analysis

The microbiome dataset used in this analysis is originally the longitudinal wild baboon gut microbiome dataset available through \texttt{microbiomeDataSets::baboongut()}, derived from the GrieneisenTS study (Grieneisen et al., 2021). The dataset contains 16,234 16S rRNA gut microbiome profiles from 585 wild baboons sampled over a 14-year period. Because the full dataset contains repeated samples from the same individuals, we construct a subject-level analysis dataset by retaining one profile per baboon, selecting the sample with the largest library size for each individual. This yields 585 independent baboon samples for downstream analysis. We use observed richness as the primary alpha-diversity endpoint and evaluate its association with host and environmental covariates, including age, sex, and season, under heterogeneous sequencing depths. This dataset provides a real-world microbiome case study for assessing whether \texttt{MBRarefy} can reduce residual library-size dependence compared with conventional single-depth rarefying while preserving biologically interpretable alpha-diversity associations.

1. Prepare Input: Per-sample count tables;

```{r eval=FALSE, echo=TRUE}
tse <- baboongut()
counts <- assay(tse, "counts")
meta <- as.data.frame(colData(tse))

subject_var <- "baboon_id"

totalReads <- colSums(counts, na.rm = TRUE)
meta$totalReads <- as.numeric(totalReads)
meta$sample_id <- colnames(counts)

keep <- !is.na(meta[[subject_var]]) &
  meta[[subject_var]] != "" &
  is.finite(meta$totalReads) &
  meta$totalReads > 0
counts <- counts[, keep, drop = FALSE]
meta <- meta[keep, , drop = FALSE]

ord <- order(meta[[subject_var]], -meta$totalReads)
meta_ord <- meta[ord, , drop = FALSE]
counts_ord <- counts[, ord, drop = FALSE]

one_idx <- !duplicated(meta_ord[[subject_var]])
dataPheno <- meta_ord[one_idx, , drop = FALSE]
rownames(dataPheno) <- dataPheno$sample_id
counts_one <- counts_ord[, one_idx, drop = FALSE]
colnames(counts_one) <- dataPheno$sample_id
```

Because \texttt{MBRarefy} is designed to support both immune-repertoire and microbiome count data, we used a common file-based input format for the baboon gut microbiome analysis. Specifically, the microbiome count table was split into separate plain-text files, with one file corresponding to one selected sample. Each file contained two columns: \texttt{seq}, giving the feature identifier, such as an ASV or sequence label, and \texttt{count}, giving the observed count of that feature in the sample. The directory specified by \texttt{InputDataDir} therefore stores the per-sample count profiles used by \texttt{multibin.rarefy.diversity()} for grid-based repeated rarefying. This format makes the microbiome analysis consistent with the package’s TCR workflow, where each sample is also represented by a separate feature-count file.

```{r eval=FALSE, echo=TRUE}
InputDataDir <- "~/Desktop/TCR/MBRarefyPaper/BaboonGut_MBRarefy_Input_one_per_baboon/"
dir.create(InputDataDir, showWarnings = FALSE, recursive = TRUE)

## remove old/stale files if any
old_files <- list.files(InputDataDir, full.names = TRUE, all.files = TRUE, no.. = TRUE)
if (length(old_files) > 0) {unlink(old_files, recursive = TRUE, force = TRUE)}

taxa_id <- rownames(counts_one)
if (is.null(taxa_id)) {taxa_id <- paste0("Taxon", seq_len(nrow(counts_one)))}

for (j in seq_len(ncol(counts_one))) {
  sample_id <- colnames(counts_one)[j]
  x <- counts_one[, j]

  df <- data.frame(seq = taxa_id, count = as.integer(x), stringsAsFactors = FALSE)
  df <- df[is.finite(df$count) & df$count > 0, , drop = FALSE]

  write.table(df, file = file.path(InputDataDir, paste0(sample_id, ".txt")),
    sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)
}
```

2. Multibin rarefying across a user-defined library size grid (with optional multiple rarefying).

```{r eval=FALSE, echo=TRUE}
SeqVar <- "seq"
CountVar <- "count"

depths <- seq(from=1e4, to=15e4, by=1e3)

alphamethods  <- c(
  "unique_seq_alpha",      # Number of unique taxa (Richness)
  "shannon_alpha",         # Shannon entropy
  "gini_simpson_alpha",    # Gini-Simpson index
  "chao1_alpha",           # Chao1 richness estimator
  "chao1_bc_alpha",        # Bias-corrected Chao1
  "pielou_alpha"           # Pielou's evenness
)

tim1 <- Sys.time()
raw.multibin.alpha.res <- multibin.rarefy.diversity(
  InputDataDir=InputDataDir,
  CountVar=CountVar,
  div.measure="alpha",
  depths=depths,
  methods=alphamethods,
  nRep=10,
  parallel=TRUE,
  nCore=10)
tim2 <- Sys.time()

tim2 - tim1
```


3. Data-adaptive library-size binning via optimal thresholds selection.

```{r}
load("../data/baboongut1.rda")

dataPheno              = baboongut1$dataPheno
raw.multibin.alpha.res = baboongut1$alpha.res
depths                 = baboongut1$depths
alphamethods           = baboongut1$methods
```


```{r}
alpha.mats <- get_alpha_metric_matrix(raw.multibin.alpha.res, metrics = alphamethods)
USC <- alpha.mats$unique_seq_alpha
dataPheno <- dataPheno[match(rownames(USC), rownames(dataPheno)), , drop = FALSE]
```

We apply the overall rarefying technique to reduce the varying library size. We select the rarefying level to be 

```{r}
Lorig <- as.numeric(dataPheno$totalReads)
Lgrid <- depths
Y <- as.matrix(USC)
overall_depths <- c(10000, 30000, 50000, 70000)

run_overall_diag <- function(x, x_name) {
  do.call(rbind, lapply(overall_depths, function(d) {
    colname <- paste0("RarefyTo", d)
    y_d <- Y[, colname]
    ok <- is.finite(y_d) & is.finite(x)
    fit <- lm(y_d[ok] ~ x[ok])
    data.frame(
      variable = x_name,
      rarefy_depth = d,
      n = sum(ok),
      correlation = cor(y_d[ok], x[ok], use = "complete.obs"),
      slope = coef(fit)[2],
      p_value = summary(fit)$coefficients[2, "Pr(>|t|)"],
      r_squared = summary(fit)$r.squared
    )
  }))
}

overall_diag_list <- list()
overall_diag_list$totalReads = run_overall_diag(x=dataPheno$totalReads, x_name="totalReads")
overall_diag_list$age = run_overall_diag(x=as.numeric(dataPheno$age), x_name="age")
sex_x <- as.numeric(as.factor(dataPheno$sex)) - 1
overall_diag_list$sex = run_overall_diag(x=sex_x, x_name="sex")
season_x <- as.numeric(as.factor(dataPheno$season)) - 1
overall_diag_list$season = run_overall_diag(x=season_x, x_name="sex")

overall_diag <- do.call(rbind, overall_diag_list)
overall_diag

```

- Model 1: With user-specified number of bins (``fixed-K'');

```{r}
fix_fit <- GAReg::gareg_knots(
  y         = rep(0, length(Lgrid)),
  x         = Lgrid,
  ObjFunc   = fixBinRegObj,
  minDist   = 1L,
  gaMethod  = "cptga",
  cptgactrl = cptgaControl(
    popSize      = 400,
    maxgen       = 100000,
    pchangepoint = 0.3
  ),
  fixedknots   = 5L,
  Y            = Y,
  Lorig        = Lorig,
  Z            = NULL,
  weight       = "count",
  min_subjects = 20L,
  use_logL     = TRUE
)

summary(fix_fit)
best_idx_fix <- as.integer(fix_fit@bestsol)

best_knots_fix <- Lgrid[best_idx_fix]
depths_op_fix <- sort(unique(c(min(Lgrid), best_knots_fix)))

best_idx_fix
best_knots_fix
depths_op_fix

```





```{r, fig.width=10, fig.height=7}
res_fix <- plotMBRarefy(
  Lorig = Lorig,
  Lgrid = Lgrid,
  Y = Y,
  fit = fix_fit,
  y_mode = "bin_lower_bound",
  title_prefix = "Select cutpoints to reduce depth-diversity dependence (Fixed-K)",
  y_label = "Observed richness at bin lower bound",
  show_missing_as_rug = FALSE
)

res_fix$plot
```

- Model 2: With estimated number of bins (``vary-K'');

```{r}
var_fit <- GAReg::gareg_knots(
  y         = rep(0, length(Lgrid)),
  x         = Lgrid,
  ObjFunc   = varBinRegObj,
  gaMethod  = "cptga",
  minDist   = 3L,
  cptgactrl = cptgaControl(
    popSize      = 400,
    maxgen       = 100000,
    pchangepoint = 0.3
  ),
  Y            = Y,
  Lorig        = Lorig,
  Z            = NULL,
  weight       = "count",
  min_subjects = 20L,
  use_logL     = TRUE
)

summary(var_fit)

best_idx_var <- as.integer(var_fit@bestsol)

best_knots_var <- Lgrid[best_idx_var]
depths_op_var <- sort(unique(c(min(Lgrid), best_knots_var)))

best_idx_var
best_knots_var
depths_op_var

```




```{r, fig.width=10, fig.height=7}
res_var <- plotMBRarefy(
  Lorig = Lorig,
  Lgrid = Lgrid,
  Y = Y,
  fit = var_fit,
  y_mode = "bin_lower_bound",
  title_prefix = "Select cutpoints to reduce depth-diversity dependence (Vary-K)",
  y_label = "Observed richness at bin lower bound",
  show_missing_as_rug = FALSE
)

res_var$plot
```

4. Multibin rarefying and alpha-diversity extraction given selected optimal thresholds.

We picked the ``fixed-K'' mode for illustration.

```{r}
USC_mat <- USC[,paste0("RarefyTo", depths_op_fix)]
```

5. Cross-bin association testing and Meta-analysis.

```{r}
res.test.libsize <- multibin.meta.test.alpha(
  y.alpha.mat=USC_mat, 
  xs=dataPheno$totalReads,
  totalReads=dataPheno$totalReads, 
  BinCuts=c(depths_op_fix, Inf))
res.test.libsize$Multi.bin.IVW
```

```{r}
res.test.age <- multibin.meta.test.alpha(
  y.alpha.mat=USC_mat, 
  xs=as.numeric(dataPheno$age),
  totalReads=dataPheno$totalReads, 
  BinCuts=c(depths_op_fix, Inf))
res.test.age$Multi.bin.IVW
```

```{r}
res.test.sex <- multibin.meta.test.alpha(
  y.alpha.mat=USC_mat, 
  xs=as.numeric(as.factor(dataPheno$sex))-1,
  totalReads=dataPheno$totalReads, 
  BinCuts=c(depths_op_fix, Inf))
res.test.sex$Multi.bin.IVW
```


```{r}
res.test.season <- multibin.meta.test.alpha(
  y.alpha.mat=USC_mat, 
  xs=as.numeric(as.factor(dataPheno$season))-1,
  totalReads=dataPheno$totalReads, 
  BinCuts=c(depths_op_fix, Inf))
res.test.season$Multi.bin.IVW
```






# Reference

Emerson, R. O., DeWitt, W. S., Vignali, M., Gravley, J., Hu, J., et al. (2017). *Immunosequencing identifies signatures of cytomegalovirus exposure history and HLA-mediated effects on the T cell repertoire*. Nature Genetics, 49(5), 659–665.

Grieneisen, L. E., Dasari, M., Gould, T. J., Björk, J. R., Grenier, J.-C., Yotova, V., et al. (2021). *Gut microbiome heritability is nearly universal but environmentally contingent*. Science, 373(6551), 181–186.

Li, M., Hua, X., Li, S., Wu, M. C., \& Zhao, N. (2024). *A multi-bin rarefying method for evaluating alpha diversities in TCR sequencing data*. Bioinformatics, 40(7), btae431.

Li, M., Lu, QQ., Lund, R., \& Shi, X. (2026). *Genetic Algorithms in Regression*.

Li, M. (2026). *MBRarefy: data-adaptive library-size binning and multibin rarefying for alpha diversity association analysis*. 

